---
title: "Working with R and RStudio"
subtitle: "A short code-along"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

## Console and document

Run `2 + 2` in the Console. Then run the chunk below.

```{r}
2 + 2
```

Both commands are evaluated by R. Only one is recorded in this document.

## The Console is a REPL

The Console is a **read-evaluate-print loop**:

1. R reads the expression entered at the `>` prompt.
2. R evaluates it.
3. R prints the result.
4. The `>` prompt returns for the next expression.

Type `1 +` in the Console and press Enter. The `+` prompt means that R is
waiting for the rest of an incomplete expression. Press Escape to cancel it.

In a Quarto code chunk, `Ctrl+Enter` sends the current line to the Console on
Windows.

## Question and prediction

What is the mean weight of these three recorded mice?

::: {.analysis-prompt}
Before running the calculation, predict an approximate value. The mean should
lie between the smallest and largest recorded weights.
:::

Use each recorded weight once and calculate the arithmetic mean in grams.

## Common atomic storage types

Run each expression and compare the results.

```{r}
typeof(18.2)
typeof(18L)
typeof("M01")
typeof(TRUE)
```

- Ordinary numerical measurements such as `18.2` are stored as doubles.
- The `L` in `18L` marks an integer.
- Quoted text is a character value.
- `TRUE` and `FALSE` are logical values.

Complex and raw values complete R's six atomic storage types. They are not
needed for this analysis.

## Missing values retain a type

Run each expression. A bare `NA` is logical; the suffix records another
underlying type.

```{r}
typeof(NA)
typeof(NA_integer_)
typeof(NA_real_)
typeof(NA_complex_)
typeof(NA_character_)
```

Use a predicate to find missing values:

```{r}
is.na(c("M01", NA_character_, "NA"))
```

`NA_character_` is a missing character value. `"NA"` is two letters stored as
text. Raw vectors are the one atomic type without a missing-value constant.

## Objects and vectors

Before running this chunk, predict the value returned by `mean()`.

```{r}
weights_g <- c(18.2, 19.5, 20.1)
typeof(weights_g)
length(weights_g)
mean(weights_g)
```

- `weights_g` names an object.
- `c()` creates a vector whose values have one underlying type.
- `length()` and `mean()` are functions.
- The values inside `c()` are arguments.

Run `?mean` in the Console to open its help page.

## A small tibble

```{r}
#| message: false

library(tidyverse)

measurements <- tibble(
  mouse_id = c("M01", "M02", "M03"),
  weight_g = weights_g
)

measurements
glimpse(measurements)
typeof(measurements)
class(measurements)
```

::: {.analysis-prompt}
Identify the package, object, functions, arguments and column names in this
chunk. Which column is character, and which is double?
:::

Each tibble column is a vector. Columns in the same tibble have the same
length. `typeof()` describes how an object is stored; `class()` helps determine
how functions interpret it.

## Missing, undefined, infinite or absent?

```{r}
special_values <- c(18.2, NA_real_, NaN, Inf, -Inf)

tibble(
  value = special_values,
  missing = is.na(special_values),
  undefined = is.nan(special_values),
  infinite = is.infinite(special_values),
  finite = is.finite(special_values)
)
```

- `NA` records an unknown or missing value.
- `NaN` is an undefined numerical result; it is also detected by `is.na()`.
- `Inf` and `-Inf` are infinite numerical results, not missing values.
- `NULL` represents absence rather than a missing observation.

```{r}
typeof(NULL)
length(NULL)
is.null(NULL)
```

## Missingness is an analytical decision

```{r}
weights_with_missing <- c(18.2, NA_real_, 20.1)
mean(weights_with_missing)
mean(weights_with_missing, na.rm = TRUE)
```

::: {.analysis-prompt}
How many observations contribute to the second mean? What would justify
excluding the missing value from this calculation?
:::

`na.rm = TRUE` changes the calculation. It should record a scientific decision,
not serve as routine cleanup.

## Result and check

```{r}
length(weights_g)
range(weights_g)
mean(weights_g)
```

::: {.analysis-prompt}
Report the mean with its unit. Confirm that three values were used and that the
mean lies within their observed range.
:::

This result describes these three records. It does not establish a population
average or a treatment effect.

## Read an error

Type this command in the Console:

```r
mean(weight_g)
```

Read the error before changing the command. Which object exists, and where can
you confirm its name?

`weight_g` is not a standalone object. It is also the name of a column inside
`measurements`. Use `$` to state where that column lives:

```{r}
names(measurements)
measurements$weight_g
mean(measurements$weight_g)
```

## Read a warning

Predict the type of `mixed_weights`, then run the chunk.

```{r}
mixed_weights <- c(18.2, 19.5, "20.1")
typeof(mixed_weights)
mean(mixed_weights)
```

The quoted value makes the whole atomic vector character. `mean()` completes,
but warns that its input is not numeric or logical and returns `NA`. Read a
warning before deciding whether its output is usable.

## Restart and render

1. Save this document.
2. Select **Session > Restart R**.
3. Check that the Environment is empty.
4. Select **Render**.

If the document renders, the code it needs is recorded in the document and can
be run from a clean session.

## What persisted?

::: {.analysis-prompt}
After restarting R, what disappeared? What remained on disk? Which one is the
durable record of this work?
:::
