---
title: "Expression filtering and transformation"
subtitle: "Reason about what remains and how values are represented"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

expression_counts <- matrix(
  c(
    0, 0, 0, 0,
    1, 0, 2, 1,
    8, 10, 31, 28,
    45, 52, 48, 55,
    120, 110, 420, 390,
    900, 1050, 980, 1120
  ),
  nrow = 6,
  byrow = TRUE,
  dimnames = list(
    paste0("gene_", LETTERS[1:6]),
    c("control_1", "control_2", "treated_1", "treated_2")
  )
)

sample_info <- tibble::tribble(
  ~sample_id, ~treatment,
  "control_1", "control",
  "control_2", "control",
  "treated_1", "treated",
  "treated_2", "treated"
)
```

## Situation

Before comparing treatment groups, remove genes with almost no observed signal
and transform the retained counts so large values do not dominate inspection.

## Biological question

::: {.analysis-prompt}
State the biological comparison the expression data should support.
:::

## Inspect the representation

::: {.analysis-prompt}
Inspect dimensions, row and column labels, representative values, summaries,
and the link between matrix columns and `sample_info`.
:::

```{r}
expression_counts
sample_info
```

## Commit before changing the data

::: {.analysis-prompt}
Use this filtering rule: retain a gene when its count is at least 10 in at
least two samples. Before applying it, list the genes you expect to retain,
predict the resulting dimensions, and identify any boundary values.
:::

Filtering changes what remains. State whether this rule operates across genes
or samples and why that axis matches the scientific goal.

## Filter and checkpoint

```{r}
# Apply the filter, then compare dimensions, gene names, and representative rows.
```

::: {.analysis-prompt}
Compare the observed retained genes and dimensions with your prediction. Explain
any disagreement before continuing.
:::

## Transform and checkpoint

Use `log2(count + 1)` on the retained matrix. Transformation changes how
retained values are represented.

::: {.analysis-prompt}
Before applying it, predict which differences will appear compressed and which
ordering relationships should remain unchanged.
:::

Normalization affects how samples can be compared. Discuss it conceptually
here; do not implement a normalization method in this exercise.

```{r}
# Apply the transformation, then compare values, summaries, and distributions.
```

::: {.analysis-prompt}
Compare representative values, summaries, distributions, and sample labels
before and after transformation.
:::

## Compare strategies

::: {.analysis-prompt}
Ask AI for two plausible preprocessing strategies. Record both, test each on
the same fixture, and identify evidence that distinguishes them.
:::

<!-- Do not choose a strategy because it runs. Compare its consequences. -->

## Result

::: {.analysis-prompt}
Show the before-and-after dimensions, representative values, and visual pattern.
:::

## Interpretation

::: {.analysis-prompt}
Explain how the representation choices affect the biological comparison.
:::

## Why I trust this

::: {.analysis-prompt}
Point to checks that distinguish filtering from transformation and reveal a
misaligned sample or implausible value.
:::

## Optional extensions

### Change the filtering assumption

Require a count of at least 10 in all four samples. Predict which genes change,
apply the revised rule, and compare the biological consequence.

### Transfer the workflow to airway

The small matrix makes every consequence visible. Once the course `airway`
dependencies are installed, repeat the same checks on the RNA-seq count matrix.

```{r}
#| eval: false

data(airway, package = "airway")
airway_counts <- SummarizedExperiment::assay(airway)
dim(airway_counts)
rownames(airway_counts)[1:6]
colnames(airway_counts)
```
