---
title: "Expression filtering and transformation"
subtitle: "Reason about what remains and how values are represented"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

expression_counts <- matrix(
  c(
    0, 0, 0, 0,
    1, 0, 2, 1,
    8, 10, 31, 28,
    45, 52, 48, 55,
    120, 110, 420, 390,
    900, 1050, 980, 1120
  ),
  nrow = 6,
  byrow = TRUE,
  dimnames = list(
    paste0("gene_", LETTERS[1:6]),
    c("control_1", "control_2", "treated_1", "treated_2")
  )
)

sample_info <- tibble::tribble(
  ~sample_id, ~treatment,
  "control_1", "control",
  "control_2", "control",
  "treated_1", "treated",
  "treated_2", "treated"
)
```

## Biological question

::: {.analysis-prompt}
State the biological comparison the expression data should support.
:::

## Prediction

::: {.analysis-prompt}
Predict how filtering or transformation could change the visible pattern.
:::

## Inspect the representation

::: {.analysis-prompt}
Inspect dimensions, row and column labels, representative values, summaries,
and the link between matrix columns and `sample_info`.
:::

```{r}
expression_counts
sample_info
```

## Bridge to airway

The small matrix makes every consequence visible. Once the course `airway`
dependencies are installed, repeat the same checks on the RNA-seq count matrix.

```{r}
#| eval: false

data(airway, package = "airway")
airway_counts <- SummarizedExperiment::assay(airway)
dim(airway_counts)
rownames(airway_counts)[1:6]
colnames(airway_counts)
```

## Filtering rule

::: {.analysis-prompt}
Specify which genes should remain and why. Filtering changes what remains.
:::

<!-- State the threshold, how many samples must pass it, and boundary behavior. -->

## Filter and checkpoint

```{r}
# Apply the filter, then compare dimensions, gene names, and representative rows.
```

## Transformation rule

::: {.analysis-prompt}
Specify how retained values should change. Transformation changes how retained values are represented.
:::

<!-- State the transformation, any offset, and what problem it is intended to address. -->

Normalization affects how samples can be compared. Discuss it conceptually
here; do not implement a normalization method in this exercise.

## Transform and checkpoint

```{r}
# Apply the transformation, then compare values, summaries, and distributions.
```

## Compare strategies

::: {.analysis-prompt}
Ask AI for two plausible preprocessing strategies. Record both, test each on
the same fixture, and identify evidence that distinguishes them.
:::

<!-- Do not choose a strategy because it runs. Compare its consequences. -->

## Result

::: {.analysis-prompt}
Show the before-and-after dimensions, representative values, and visual pattern.
:::

## Interpretation

::: {.analysis-prompt}
Explain how the representation choices affect the biological comparison.
:::

## Why I trust this

::: {.analysis-prompt}
Point to checks that distinguish filtering from transformation and reveal a
misaligned sample or implausible value.
:::
