---
title: "Getting started with BIOSCI504"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with BIOSCI504}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`BIOSCI504` installs the R packages used in Days 1--4 and provides the course
data, projects, and exercise templates.

## First installation on Windows

Use this route if `BIOSCI504` is not already installed. It installs the
released Windows binary and does not require Rtools.

```{r install, eval = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("pak")
pak::pkg_install(
  "BIOSCI504=url::https://github.com/cohmathonc/BIOSCI504/releases/latest/download/BIOSCI504-windows.zip",
  upgrade = FALSE
)
```

## Updating an existing installation on Windows

Use this route if `BIOSCI504` is already installed. Save your work and restart
R before updating the package. Do not load `BIOSCI504` in the new session.
Then run:

```{r update, eval = FALSE}
pak::pkg_install(
  "BIOSCI504=url::https://github.com/cohmathonc/BIOSCI504/releases/latest/download/BIOSCI504-windows.zip?reinstall",
  upgrade = FALSE,
  ask = FALSE
)
```

Restart R again after the installation finishes.

## Check your setup

Run the setup check before the first class:

```{r check-setup, eval = FALSE}
BIOSCI504::check_setup()
```

The check reports every problem it finds and suggests a fix. RStudio may be
reported as a warning when the check is run from a terminal.

## Create a course project

Choose a folder that is easy to find:

```{r create-project, eval = FALSE}
BIOSCI504::create_course_project("~/BIOSCI504")
```

Open `~/BIOSCI504/BIOSCI504.Rproj` in RStudio before continuing. This keeps
course files together, applies the course repository settings, and makes
relative paths predictable.

## Copy an exercise

From the course project, copy the Day 2 template:

```{r copy-template, eval = FALSE}
BIOSCI504::copy_template("tabular-comparison")
```

The template is copied to `exercises/tabular-comparison`. Existing work is not
replaced unless you explicitly request it. Updating the package does not
change an exercise that has already been copied.

## Work with course data

Load the package and tidyverse, then inspect the bundled mouse trial:

```{r load-data, message = FALSE}
library(BIOSCI504)
library(tidyverse)

data(mouse_trial)
glimpse(mouse_trial)
```

A quick plot is often the best next check:

```{r plot-data, fig.width = 6, fig.height = 4}
ggplot(
  mouse_trial,
  aes(
    x = baseline_weight_g,
    y = final_weight_g,
    colour = treatment
  )
) +
  geom_point() +
  labs(
    x = "Baseline weight (g)",
    y = "Final weight (g)",
    colour = "Treatment"
  )
```
