---
title: "Tabular comparison"
subtitle: "From a biological question to a trustworthy interpretation"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
library(tidyverse)
data(mouse_trial)
```

## Situation

A recovery study tests whether a nutritional treatment increases weight gain
relative to a control. Each mouse was weighed before and after treatment. Mice
shared cages, and measurements were collected in batches.

## Question and prediction

::: {.analysis-prompt}
State whether the treatment groups changed weight differently. Predict the
direction of the difference and explain the biological reasoning.
:::

<!-- Commit to an expectation before seeing the group comparison. -->

## How the data arose

::: {.analysis-prompt}
Describe how treatment, starting weight, cage, batch, and measurement problems
could contribute to the observed final weights.
:::

<!-- Keep this as a short verbal model. It should guide what you inspect. -->

## Analysis specification

::: {.analysis-prompt}
Plan two comparisons: final weight by treatment, then weight change by
treatment. Specify the unit of observation and important data-handling choices.
:::

<!--
Include exclusions, missing-value handling, summary or model, and planned
figure. This may support a consequential AI prompt, but it must stand on its own
as a clear analytical plan.
-->

## Data inspection

::: {.analysis-prompt}
Inspect the data before transforming or plotting it. Record anything that could
change the analysis.
:::

<!--
Check the dimensions, variable names and types, identifiers, group sizes,
missing values, ranges, and units. Do not silently repair unexpected values.
-->

```{r}
# Start by inspecting the structure, dimensions, names, and missingness.
```

## Explain or debug with AI

::: {.analysis-prompt}
Choose one line of inspection code, unexpected output, or error. Ask AI to
explain it without generating the analysis. Record the explanation, then test
it against the data or a small code change.
:::

<!-- Preserve the original code, output, or error beside the explanation. -->

## Compare final weight

::: {.analysis-prompt}
Compare final weight between treatment groups. Record what this comparison can
and cannot show about change during the study.
:::

```{r}
# Summarize and visualize final weight by treatment.
```

### Checkpoint

::: {.analysis-prompt}
Check group sizes, missing values, ranges, and whether individual observations
support the summary.
:::

## Derive and compare weight change

::: {.analysis-prompt}
Create `weight_change_g`, inspect it, and compare it between treatment groups.
Explain why this answers a different question from final weight.
:::

```{r}
# Derive weight change, inspect it, then summarize and visualize it by treatment.
```

### Checkpoint

::: {.analysis-prompt}
Record missing values introduced, implausible changes, group summaries, and any
observation that strongly influences the comparison.
:::

## Final verification

::: {.analysis-prompt}
Show the checks that would reveal a wrong unit, duplicated mouse, unexpected
missingness, or a mismatch between the planned and executed comparison.
:::

<!--
Use evidence rather than reassurance. Useful checkpoints include dimensions,
unique identifiers, group counts, missing-value counts, ranges, and intermediate
summaries.
-->

## Result

::: {.analysis-prompt}
Report the main numerical and visual result without interpreting it yet.
:::

<!--
Include the size and direction of the observed difference and the uncertainty
needed for the planned analysis. Ensure every figure can be read independently.
-->

## Interpretation

::: {.analysis-prompt}
Explain what the result means for the biological question.
:::

<!--
Separate what the data show from possible biological explanations. State the
limits of the design and avoid causal language the study cannot support.
-->

## Why I trust this

::: {.analysis-prompt}
Summarize the evidence that makes the result credible and any remaining doubt.
:::

<!--
Point to specific checks, not effort or confidence. Note assumptions that remain
untested and describe what additional evidence would change the conclusion.
-->

## Optional extension

::: {.analysis-prompt}
Choose one design feature—sex, cage, or batch—and show whether the treatment
pattern is consistent across it. Explain what this check can and cannot resolve.
:::
