---
title: "Tabular comparison"
subtitle: "From a biological question to a trustworthy interpretation"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
data(mouse_trial)
```

## Biological question

::: {.analysis-prompt}
State the biological question in one sentence.
:::

<!--
Name the population, treatment or exposure, and outcome. Keep the question
answerable with the available data.
-->

## Prediction

::: {.analysis-prompt}
Describe the pattern you expect to observe and why.
:::

<!--
Make the direction of the expected difference explicit. Connect the prediction
to biological reasoning rather than to a preferred statistical result.
-->

## Analysis specification

::: {.analysis-prompt}
Specify the comparison, outcome, groups, and important data-handling choices.
:::

<!--
Write the analysis before carrying it out. Include the unit of observation,
exclusions, missing-value handling, summary or model, and planned figure. This
specification may also support a consequential AI prompt, but it must stand on
its own as a clear analytical plan.
-->

## Data inspection

::: {.analysis-prompt}
Inspect the data before transforming or plotting it. Record anything that could
change the analysis.
:::

<!--
Check the dimensions, variable names and types, identifiers, group sizes,
missing values, ranges, and units. Do not silently repair unexpected values.
-->

```{r}
# Start by inspecting the structure, dimensions, names, and missingness.
```

## Analysis code

::: {.analysis-prompt}
Carry out the specified comparison with readable, reproducible code.
:::

<!--
Keep transformations close to the analysis that uses them. Prefer explicit
names and small steps that another student can inspect.
-->

## Verification checkpoints

::: {.analysis-prompt}
Show the checks that would reveal a wrong unit, duplicated observation,
unexpected missingness, or implausible result.
:::

<!--
Use evidence rather than reassurance. Useful checkpoints include dimensions,
unique identifiers, group counts, missing-value counts, ranges, and intermediate
summaries.
-->

## Result

::: {.analysis-prompt}
Report the main numerical and visual result without interpreting it yet.
:::

<!--
Include the size and direction of the observed difference and the uncertainty
needed for the planned analysis. Ensure every figure can be read independently.
-->

## Interpretation

::: {.analysis-prompt}
Explain what the result means for the biological question.
:::

<!--
Separate what the data show from possible biological explanations. State the
limits of the design and avoid causal language the study cannot support.
-->

## Why I trust this

::: {.analysis-prompt}
Summarize the evidence that makes the result credible and any remaining doubt.
:::

<!--
Point to specific checks, not effort or confidence. Note assumptions that remain
untested and describe what additional evidence would change the conclusion.
-->
