---
title: "Explicit analytical rules"
subtitle: "From biological intent to executable conditions"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
library(tidyverse)
data(mouse_trial)

rule_cases <- tibble::tribble(
  ~mouse_id, ~treatment, ~baseline_weight_g, ~final_weight_g,
  "case_1", "vehicle", 24.0, 24.4,
  "case_2", "treatment", 25.0, 23.2,
  "case_3", "treatment", 22.5, NA_real_,
  "case_4", "treatment", 21.0, 20.0
)
```

## Situation

A study team needs a reproducible rule for identifying mouse records that
require review. The same decision must behave consistently for ordinary,
missing, and boundary values.

## Review rule

::: {.analysis-prompt}
Use this practice rule: classify a record as requiring review when final weight
is missing. Otherwise, flag a treated mouse that lost at least 1.0 g and
classify all other records as not flagged.
:::

<!-- Identify the measurements, group, threshold, and missing-value behavior. -->

## Expected classifications

::: {.analysis-prompt}
Classify each practice case before writing code. Pay particular attention to
the missing value and the value exactly on the 1.0 g boundary.
:::

| Case | Expected classification | Reason |
|---|---|---|
| `case_1` |  |  |
| `case_2` |  |  |
| `case_3` |  |  |
| `case_4` |  |  |

## Executable specification

::: {.analysis-prompt}
Translate the rule into ordered, explicit comparisons joined with `&` or `|`.
:::

<!-- Check missing values before applying the numeric boundary. -->

## Constrained AI generation

::: {.analysis-prompt}
Give the executable specification to AI as a constrained prompt. Preserve the
prompt and generated code, but test the code before accepting it.
:::

<!-- Ask for one operation and name the permitted inputs, output, and checks. -->

## Small-case test

The four cases include an ordinary value, a clear change, a missing value, and
a treated mouse exactly at the boundary.

```{r}
rule_cases
```

```{r}
# Apply the rule to rule_cases. Compare the result with your expected outcomes.
```

## Scale the operation

::: {.analysis-prompt}
Use `filter()` to retain records that meet the numeric flag condition. Track
records with missing final weight separately rather than dropping them silently.
:::

```{r}
# Inspect mouse_trial, derive any needed variables, then apply the rule.
```

### Checkpoint

::: {.analysis-prompt}
Compare dimensions, flagged IDs, treatment counts, and representative rows
before and after the operation.
:::

## Conditional outcome

::: {.analysis-prompt}
Use `if_else()` or `case_when()` to label records as flagged, not flagged, or
requiring review. Check that these labels reproduce the expected practice-case
classifications.
:::

## Custom function

::: {.analysis-prompt}
Wrap one well-tested rule in a small function with a descriptive name.
:::

<!-- Keep the function focused. Its inputs and output should be easy to explain. -->

## Explicit iteration

::: {.analysis-prompt}
Apply the function across a short, visible set of thresholds using
`purrr::map()`. Predict what one threshold change should do before iterating;
the missing-value classification should remain unchanged.
:::

<!-- First write down what one iteration receives and returns. -->

## Final verification

::: {.analysis-prompt}
Show that the small-case expectations, scaled classifications, conditional
labels, function output, and iterated results agree.
:::

## Result

::: {.analysis-prompt}
Report what the rule retained or classified without interpreting it yet.
:::

## Interpretation

::: {.analysis-prompt}
Explain the biological meaning of the result and the consequence of the rule.
:::

## Why I trust this

::: {.analysis-prompt}
Point to the small-case test and scaled checkpoints that could reveal an error.
:::

## Optional extensions

### Change the boundary

Suppose the policy changes from “at least 1.0 g” to “more than 1.0 g.” Predict
which cases change, update the rule, and rerun the same checks.

### Transfer the workflow to airway

`airway` is the course anchor for high-dimensional data. Once its instructor
dependencies are installed, extract its sample metadata and apply the same
inspect → specify → test → scale workflow.

```{r}
#| eval: false

data(airway, package = "airway")
airway_samples <- SummarizedExperiment::colData(airway) |>
  as.data.frame() |>
  tibble::rownames_to_column("sample_id") |>
  tibble::as_tibble()
```
