---
title: "Logical operations in the mouse trial"
subtitle: "Comparisons, conditions, and filtering"
brand: _brand.yml
format:
  html:
    theme:
      - cosmo
      - brand
    toc: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
library(tidyverse)
data(mouse_trial)

mouse_trial <- mouse_trial |>
  as_tibble() |>
  mutate(weight_change_g = final_weight_g - baseline_weight_g)
```

## Start

Work inside your BIOSCI504 course Project.

1. Render this untouched document once.
2. Complete one section at a time and run its code chunk.
3. Keep `mouse_id` beside your logical results.
4. Use each checkpoint before continuing.
5. Restart R and render the completed document.

## Inspect the mouse trial

Each row represents one mouse. Inspect the columns needed for this exercise.

```{r}
mouse_trial |>
  select(
    mouse_id,
    treatment,
    baseline_weight_g,
    final_weight_g,
    weight_change_g
  ) |>
  slice_head(n = 8)
```

The question for this exercise is:

> Which treated mice gained at least 3 g?

The 3 g threshold is used to practise explicit logical operations. It is not a
biological response category.

## Identify missing measurements

A change in weight requires both the baseline and final measurements.

::: {.analysis-prompt}
Create `trial_checks` from `mouse_trial`. Add `missing_required_weight`, which is
`TRUE` when either `baseline_weight_g` or `final_weight_g` is missing.

Use `is.na()` and logical OR (`|`).
:::

```{r}
# Create trial_checks here.

# Then inspect mouse_id, both weights, weight_change_g,
# and missing_required_weight.
```

Your checkpoint is two incomplete records. Their `weight_change_g` values
should also be `NA`.

::: {.analysis-prompt}
Record the two incomplete `mouse_id` values.
:::

**Response:**

## Keep complete records

Logical NOT (`!`) reverses `TRUE` and `FALSE`. Use it to retain records that
are not missing a required weight.

::: {.analysis-prompt}
Create `complete_mouse_records` from `trial_checks` by filtering on
`!missing_required_weight`. Print the number of rows.
:::

```{r}
# Create complete_mouse_records here.

# Then use summarise(rows = n()) as a checkpoint.
```

Your checkpoint is 46 complete records.

## Compare treatment labels

The equality operator (`==`) compares each stored value with the value on its
right. The not-equal operator (`!=`) tests whether two values differ.

::: {.analysis-prompt}
Create `mouse_conditions` from `complete_mouse_records`. Add
`received_treatment`, which compares `treatment` with the stored label
`"treatment"`.
:::

```{r}
# Create mouse_conditions here.

# Then count treatment and received_treatment.
```

Check that the logical values agree with the treatment labels before
continuing.

## Compare weight change

“At least 3 g” means greater than or equal to 3 (`>= 3`). A gain of exactly
3 g is included. The related operators are `>`, `<`, and `<=`.

::: {.analysis-prompt}
Add `gained_at_least_3_g` to `mouse_conditions`. Compare `weight_change_g`
with 3, then count the resulting logical values.
:::

```{r}
# Add gained_at_least_3_g here.

# Then count gained_at_least_3_g.
```

Your checkpoint is five `TRUE` values.

## Make one decision

`if` uses one logical value to choose which branch runs.

::: {.analysis-prompt}
Complete the supplied `if`/`else` example. With
`current_weight_change_g <- 3`, it should return `"include"`. Change the value
to 2.9 and confirm that it returns `"exclude"`.
:::

```{r}
current_weight_change_g <- 3

if (current_weight_change_g >= 3) {
  # Return "include".
} else {
  # Return "exclude".
}
```

## Combine the conditions

Logical AND (`&`) returns `TRUE` only when both component conditions are
`TRUE`.

::: {.analysis-prompt}
Add `meets_both_conditions` to `mouse_conditions` by combining
`received_treatment` and `gained_at_least_3_g` with `&`.

Keep the two component conditions beside the combined result when you inspect
the table.
:::

```{r}
# Add meets_both_conditions here.

# Then inspect mouse_id, treatment, weight_change_g,
# received_treatment, gained_at_least_3_g, and meets_both_conditions.
```

Your checkpoint is five records for which `meets_both_conditions` is `TRUE`.

## Filter the selected mice

`filter()` retains rows where its condition is `TRUE`. Rows with `FALSE` or
`NA` are not retained.

::: {.analysis-prompt}
Create `treated_mice_meeting_gain_threshold` by filtering
`mouse_conditions` on `meets_both_conditions`. Retain `mouse_id`, `treatment`,
and `weight_change_g`.
:::

```{r}
# Create and print treated_mice_meeting_gain_threshold here.

# Then check rows and distinct mouse_id values with summarise().
```

Your checkpoint is five rows and five distinct mouse IDs.

::: {.analysis-prompt}
Record the five selected `mouse_id` values. Explain how the row count relates
to the preceding logical count.
:::

**Response:**

## Give the filter a reusable name

The familiar filter can be placed inside a function. The table and minimum
gain become the function inputs. Inside the function, `dplyr::filter()` names
the tidyverse function explicitly.

::: {.analysis-prompt}
Complete the two comparisons inside
`filter_treated_mice_by_weight_gain()`. Use the function once with `mouse_trial` and
`minimum_weight_gain_g = 3`.
:::

```{r}
filter_treated_mice_by_weight_gain <- function(
  mouse_data,
  minimum_weight_gain_g
) {
  mouse_data |>
    dplyr::filter(
      # Compare treatment with "treatment",
      # compare weight_change_g with minimum_weight_gain_g
    )
}

# Call the function and inspect mouse_id, treatment, and weight_change_g.
```

The function call should reproduce the same five mouse IDs.

## Repeat the filter

This supplied loop calls the same function at minimum gains of 2, 3, and 4 g.
Run it after the function returns the expected five-row result. Remove the

```text
#| eval: false
```

line before you run or render this chunk.

```{r}
#| eval: false

for (minimum_weight_gain_g in c(2, 3, 4)) {
  selected_mice <- filter_treated_mice_by_weight_gain(
    mouse_trial,
    minimum_weight_gain_g
  )

  print(nrow(selected_mice))
}
```

The three row counts should be 21, 5, and 1, in that order.

::: {.analysis-prompt}
Why does the number of selected mice decrease as the minimum gain increases?
:::

**Response:**

## Choose a continuation

Once the core checkpoints work, choose one of these continuations. **Consolidate**
provides another guided pass through the logical operations. **Extend** asks you
to adapt the function and examine the experimental design.

### Consolidate: change the question

> Which complete control mice gained at least 2 g?

This uses the same sequence with a different treatment label and threshold.
Complete the two comparisons, remove the `#| eval: false` line, and run the
chunk.

```{r}
#| eval: false

control_mice_gaining_at_least_2_g <- complete_mouse_records |>
  mutate(
    is_control = treatment == _____,
    gained_at_least_2_g = weight_change_g _____ 2
  ) |>
  filter(is_control & gained_at_least_2_g) |>
  select(mouse_id, treatment, weight_change_g)

control_mice_gaining_at_least_2_g
```

Your checkpoint is one row: `M048`. Explain why this mouse did not meet the
earlier 3 g condition.

**Response:**

### Extend: generalise and check

Create `filter_mice_by_treatment_and_gain()` by adapting
`filter_treated_mice_by_weight_gain()`. Make the treatment label an additional
input. The new function has the arguments `mouse_data`, `treatment_label`, and
`minimum_weight_gain_g`.

Use these calls as checks:

```{r}
#| eval: false

# Define filter_mice_by_treatment_and_gain() here.

filter_mice_by_treatment_and_gain(
  mouse_trial,
  treatment_label = "treatment",
  minimum_weight_gain_g = 3
)

filter_mice_by_treatment_and_gain(
  mouse_trial,
  treatment_label = "control",
  minimum_weight_gain_g = 2
)
```

The calls should return five rows and one row, respectively.

Finally, inspect whether the five mice from the original question all came
from one batch or cage. Remove the `#| eval: false` line after completing the
core exercise:

```{r}
#| echo: true
#| eval: false
mouse_conditions |>
  filter(meets_both_conditions) |>
  count(batch, cage)
```

They span three batches and three cages. If all selected mice came from one
batch or cage, that pattern would need checking before biological
interpretation.

## Finish

1. Save `analysis.qmd`.
2. Restart R: **Session > Restart R**.
3. Render the document.
4. Open `analysis.html` and confirm that it contains:
   - the two incomplete mouse IDs;
   - 46 complete records;
   - the five selected mouse IDs; and
   - threshold counts of 21, 5, and 1.

If you completed a continuation, confirm that its checkpoint also appears.

A successful render after restarting confirms that the document creates the
objects it uses.

## Further reading

- [R comparisons](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Comparison.html)
- [R logical operators](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html)
- [`is.na()` and missing values](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html)
- [`dplyr::filter()`](https://dplyr.tidyverse.org/reference/filter.html)
- [Software Carpentry: Functions](https://swcarpentry.github.io/r-novice-gapminder/instructor/10-functions.html)
- [Software Carpentry: Repeating operations](https://swcarpentry.github.io/r-novice-gapminder/instructor/07-control-flow.html#repeating-operations)
